---
title: "DATA PREPARATION"
author: "TEJO SAI KRISHNA KOTTE"
date: "10/15/23"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
```

# Importing the datasets
```{r Importing data,echo=TRUE}
# Importing the two csv files
data1<-read.csv('survey1.csv')
data2<-read.csv('survey2.csv')
dim(data1)
dim(data2)
```

# Combining the datasets
```{r Combining data,echo=TRUE}
# Combining the two dataframes to a single dataframe called survey
survey <- rbind(data1, data2)
dim(survey)
```

By comparing the dimensions we can see that we have all the rows and columns combined successfully

# Data Cleaning
```{r Data Cleaning1,echo=TRUE}
# Lets look for some missing data first
missing_data <- is.na(survey)

# Summarize missing values
summary_missing <- colSums(missing_data)

# Print the summary
print(summary_missing)
```

We can see that we have some columns that have missing values. We will first drop the comments,state,work_interfere columns because they have almost 90,50,25 percent of the missing values respectively. This makes no sense to analyse these columns

```{r Data Cleaning2,echo=TRUE}
# droping the columns
survey <- select(survey, -comments, -state,-work_interfere,-mental_health_consequence,-phys_health_consequence,-coworkers,-supervisor,mental_health_interview,-phys_health_interview,-mental_vs_physical,-obs_consequence)
dim(survey)
head(survey)
```

we can see that we have successfully dropped the columns.
We now drop the rows that have fewer missing values in the columns like Age,self_employed,family_history,treatment,tech_company
```{r Data Cleaning3,echo=TRUE}
# Remove rows with missing values in Age,self_employed,family_history,treatment,tech_company columns
survey <- na.omit(survey, cols = "Age","self_employed","family_history","treatment","tech_company")
dim(survey)
```
We have successfully removed all the NA values. Lets recheck once.

```{r Data Cleaning4,echo=TRUE}
# Lets look for some missing data first
missing_data <- is.na(survey)

# Summarize missing values
summary_missing <- colSums(missing_data)

# Print the summary
print(summary_missing)
```


```{r Data Cleaning5,echo=TRUE}
# Reordering the indexes
survey <- data.frame(survey, row.names = NULL)
dim(survey)
head(survey)
tail(survey)
```
we have successfully reordered the indexes of our dataframe.
We can see that we have successfully removed all the null  values while keeping most of our data. Now its time to check the structure of our data.
```{r Data Cleaning6,echo=TRUE}
str(survey)
```
The structure looks good except for one column (TimeStamp).We will take a look at the summary of the data first.

```{r Data Cleaning7,echo=TRUE}
summary(survey)
```
From the summary of Age column above we can see that the minimum value in the Age column is -1726. We now will clean the Age column as Age cannot be negative. And also the mean value and max value are also abnormal for Age column.
```{r Data Cleaning8,echo=TRUE}
# Plotting using ggplot2
ggplot(data = survey, aes(x = seq_along(Age), y = Age)) +
  geom_line() +
  labs(x = "Index", y = "Age", title = "Age Trend")

# Remove rows with empty Timestamp cells
survey <- survey %>%
  filter(Age != "")
# Filtering the Age column
survey <- survey %>%
  filter(Age > 0, Age < 100)
dim(survey)
```
As you can see there are some outliers in the Age column.We filtered out those using the filter method.
```{r Data Cleaning9,echo=TRUE}
#Summary on Age column
summary(survey$Age)
```
Now the Age column looks good.
We will clean the Timestamp column.
```{r Data Cleaning10,echo=TRUE}
# Remove rows with empty Timestamp cells
survey <- survey %>%
  filter(Timestamp != "")
dim(survey)
```
In the Timestamp column, the data structure is not consistent. Some of the records have "Wednesday, August 27, 2014" date format. But most of the records have "9/9/14 13:49" this format.We shall remove the records with the latter structure.

```{r Data Cleaning11,echo=TRUE}
# Remove rows bad structure of Timestamp
survey <- survey %>%
  filter(Timestamp != "Wednesday, August 27, 2014")
dim(survey)
```

The Timestamp column is of char type we will now convert it into timestamp type
```{r Data Cleaning12,echo=TRUE}
# Convert "Timestamp" to datetime type using POSIXct.
survey$Timestamp <- as.POSIXct(survey$Timestamp, format = "%m/%d/%y %H:%M")
str(survey)
```
You can see that we changed the format of our Timestamp label.
Lets move on to the gender column.
```{r Data Cleaning13,echo=TRUE}
# Remove rows with empty Gender cells
survey <- survey %>%
  filter(Gender != "")
dim(survey)

# Create a bar plot using base R
gender_counts <- table(survey$Gender)
print(gender_counts)
barplot(gender_counts, main = "Gender Distribution", xlab = "Gender", ylab = "Count")
```
We will simply have three categories namely Male, Female and Other.
```{r Data Cleaning14,echo=TRUE}
# Define a function to group genders
group_genders <- function(gender) {
  gender <- tolower(gender)  # Convert to lowercase
  if (gender %in% c("male", "m", "man", "make", "cis male", "cis man")) return("Male")
  if (gender %in% c("female", "f", "woman", "cis female", "cis woman")) return("Female")
  return("Other")
}

# Apply the grouping function to the Gender column
survey$GroupedGender <- sapply(survey$Gender, group_genders)

# Create a bar plot using ggplot2
ggplot(data = survey, aes(x = GroupedGender)) +
  geom_bar() +
  labs(x = "Gender Group", y = "Count", title = "Gender Distribution")

# Drop the old Gender column
survey <- select(survey, -Gender)
dim(survey)
```

We will now move on to the Country column.
```{r Data Cleaning15, echo=TRUE}
# Remove rows with empty Gender cells
survey <- survey %>%
  filter(Country != "")
dim(survey)
# We will plot and see the unique countries
country_counts <- table(survey$Country)
print(country_counts)
barplot(country_counts, main = "Country Distribution", xlab = "Country", ylab = "Count")
```

The country column looks good.The no_employees column has many inconsistent values. We will drop it.
```{r Data Cleaning16, echo=TRUE}
# Drop no_employees column
survey <- select(survey, columns=-no_employees)

```
Now we shall move on to the self_employed column.
```{r Data Cleaning17, echo=TRUE}
# Drop the rows with empty cells
survey <- survey %>%
  filter(self_employed != "")
dim(survey)
```
we have successfully removed 4 rows.
Now we shall look into the values in the self_employed column.
```{r Dara Cleaning18,echo=TRUE}
# Get the counts of unique values in the "self_employed" column
self_employed_counts <- table(survey$self_employed)

# Print the counts
print(self_employed_counts)
```
We can see that there are two unique values with the counts above. This column looks clean.
We will now move onto the next column family History
```{r Data Cleaning19, echo=TRUE}
# Drop the rows with empty cells
survey <- survey %>%
  filter(family_history != "")
dim(survey)
```
All the rows are clean
Now we shall look into the values in the family_history column.
```{r Dara Cleaning20,echo=TRUE}
# Get the counts of unique values in the "self_employed" column
family_history_counts <- table(survey$family_history)

# Print the counts
print(family_history_counts)
```
We can see that there are two unique values with the counts above. This column looks clean.
We will now move onto the next column treatment.
```{r Data Cleaning21, echo=TRUE}
# Drop the rows with empty cells
survey <- survey %>%
  filter(treatment != "")
dim(survey)
```
All the rows are clean
Now we shall look into the values in the treatment column.
```{r Dara Cleaning22,echo=TRUE}
treatment_counts <- table(survey$treatment)

# Print the counts
print(treatment_counts)
```
We can see that we have two rows with - value. We will drop these two rows and then we will convert the N to No and Y to Yes respectively.
```{r Data Cleaning23, echo=TRUE}
# Drop the rows with '-' value
survey <- survey %>%
  filter(treatment != "-")
dim(survey)
```

```{r Data Cleaning24, echo=TRUE}

# Replace values in the "treatment" column
survey$treatment <- ifelse(survey$treatment == "N", "No", ifelse(survey$treatment == "Y", "Yes", survey$treatment))
```
We will print te unique values once again to cross check.
```{r Dara Cleaning25,echo=TRUE}
treatment_counts <- table(survey$treatment)

# Print the counts
print(treatment_counts)
```
Now the treatment column looks good. We shall move onto the next column remote_work.
```{r Data Cleaning26, echo=TRUE}
# Drop the rows with empty cells
survey <- survey %>%
  filter(remote_work != "")
dim(survey)
```
Now we shall look into the values in the remote_work column.
```{r Dara Cleaning27,echo=TRUE}
remote_work_counts <- table(survey$remote_work)

# Print the counts
print(remote_work_counts)
```

We can see that we have two rows with - value. We will drop these two rows.
```{r Data Cleaning28, echo=TRUE}
# Drop the rows with '-' value
survey <- survey %>%
  filter(remote_work != "-")
dim(survey)
```
This column is clean. we will now move onto tech_company column.
```{r Data Cleaning29, echo=TRUE}
# Drop the rows with empty cells
survey <- survey %>%
  filter(tech_company != "")
dim(survey)
```
Now we shall look into the values in the tech_company column.
```{r Dara Cleaning30,echo=TRUE}
tech_company_counts <- table(survey$tech_company)

# Print the counts
print(tech_company_counts)
```
We have one - value. we shall drop that row
```{r Data Cleaning31, echo=TRUE}
# Drop the rows with '-' value
survey <- survey %>%
  filter(tech_company != "-")
dim(survey)
```
Now we will move onto the benefits column
```{r Data Cleaning32, echo=TRUE}
# Drop the rows with empty cells
survey <- survey %>%
  filter(benefits != "")
dim(survey)
```
Now we shall look into the values in the benefits column.
```{r Dara Cleaning33,echo=TRUE}
benefits_counts <- table(survey$benefits)

# Print the counts
print(benefits_counts)
```



```{r Dara Cleaning34,echo=TRUE}
# Clean the "benefits" column

# Define a function to group benefits
group_benefits <- function(benefit) {
  benefit <- tolower(benefit)  # Convert to lowercase
  if (benefit %in% c("don't know", " not sure")) return("Not sure")
  if (benefit %in% c(" no ")) return("No")
  if (benefit %in% c("yes")) return("Yes")
  return("Other")
}

# Apply the grouping function to the Gender column
survey$benefits <- sapply(survey$benefits, group_benefits)
benefits_counts <- table(survey$benefits)

# Print the counts
print(benefits_counts)
```
We have successfully cleaned the benefits column. We will move onto the next column care_options.
Now we will move onto the benefits column
```{r Data Cleaning35, echo=TRUE}
# Drop the rows with empty cells
survey <- survey %>%
  filter(care_options != "")
dim(survey)
```
Now we shall look into the values in the care_options column.
```{r Dara Cleaning36,echo=TRUE}
care_options_counts <- table(survey$care_options)

# Print the counts
print(care_options_counts)
```
This column looks good. Moving onto wellness_program column.

```{r Data Cleaning37, echo=TRUE}
# Drop the rows with empty cells
survey <- survey %>%
  filter(wellness_program != "")
dim(survey)
```
Now we shall look into the values in the wellness_program column.
```{r Dara Cleaning38,echo=TRUE}
wellness_program_counts <- table(survey$wellness_program)

# Print the counts
print(wellness_program_counts)
```
This column looks good, we shall move onto the next column seek_help

```{r Data Cleaning39, echo=TRUE}
# Drop the rows with empty cells
survey <- survey %>%
  filter(seek_help != "")
dim(survey)
```
Now we shall look into the values in the seek_help column.
```{r Dara Cleaning40,echo=TRUE}
seek_help_counts <- table(survey$seek_help)

# Print the counts
print(seek_help_counts)
```
We can use the group_benefits function we defined above to clean this column

```{r Data Cleaning41,echo=TRUE}
survey$seek_help <- sapply(survey$seek_help, group_benefits)
seek_help_counts <- table(survey$seek_help)

# Print the counts
print(seek_help_counts)
```
We shall omit all the rows with empty cells.
```{r Data Cleaning42,echo=TRUE}
remove_empty_rows <- function(data, column_name) {
  data %>%
    filter(vars(column_name)!="")
}

survey <- remove_empty_rows(survey,anonymity)
survey <- remove_empty_rows(survey,leave)
survey <- remove_empty_rows(survey,mental_health_consequence)
survey <- remove_empty_rows(survey,phys_health_consequence)
survey <- remove_empty_rows(survey,coworkers)
survey <- remove_empty_rows(survey,supervisor)
survey <- remove_empty_rows(survey,mental_health_interview)
survey <- remove_empty_rows(survey,phys_health_interview)
survey <- remove_empty_rows(survey,mental_vs_physical)
survey <- remove_empty_rows(survey,obs_consequence)
dim(survey)
```

```{r Data Cleaning43,echo=TRUE}
counts <- table(survey$anonymity)
print(counts)

```

We can use the group_benefits function we defined above to clean this column

```{r Data Cleaning44,echo=TRUE}
survey$anonymity <- sapply(survey$anonymity, group_benefits)
counts <- table(survey$anonymity)

# Print the counts
print(counts)
```
We have cleaned all of our required columns. Now we need to save this dataframe into a csv file for later usage.
# Saving the dataframe into csv file.
```{r Data Cleaning45,echo=TRUE}
write.csv(survey, file = "survey.csv", row.names = FALSE)
```
